---
title: "Net Ecological Footprint of Countries in the World"
author: "Shirley Cheung and Tuan Nguyen"
date: "April 30, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
It currently takes 1.5 Earths to feed 7 billion people. With the exponential growth of global population, we are consuming natural resources at an alarming rate. To measure mankind's existential pressure on the environment, Mathis Wackernagel and William Rees at the University of British Columbia came up with two quantities: ecological footprint and biocapacity. The former measures the ecological assets that are required to sustain a given population. The latter tracks the productivity of the same ecological assets, accounting for waste accumulation, pollution and so on. A country is running an ecological deficit when its footprint exceeds biocapacity, and running an ecological reserve otherwise. 

This project aims to predict ecological deficit / reserve of a nation in 2016 using the following country-specific characteristics:

* Population in 2016
* GDP per capita in 2016
* The continent that the country is on 
* The total food production in 2013
* The greenhouse gas emission in 2012
* The human development index in 2015

The data on net ecological footprint, population, continent and GDP per capita was collected by Global Footprint Network and downloaded from [Kaggle](https://www.kaggle.com/footprintnetwork/ecological-footprint). This is our main source of predictors and response variables. For this data we replaced continent (e.g Middle East/Central Asia) by its abbreviation (e.g AS). Originally in character format, GDP per capita was converted into number through very simple string manipulation (dropping the *$* sign). The next predictor on the list, food production data, was downloaded from [Kaggle](https://www.kaggle.com/unitednations/global-food-agriculture-statistics/data), by courtesy of the [United Nation](http://data.un.org/Explorer.aspx). This dataset spans between 1990 and 2013. We took the latest data (2013) and created a new variable by summing up food production across different categories. The Greenhouse Gas Emissions data by Country is collected by [World Resources Institue](http://www.wri.org/resources/data-sets/cait-country-greenhouse-gas-emissions-data). Similar to the food production data, we picked the most recent year (2012) of CO2 emission for each country. Last but not least is the Human Development Index data in 2015, by courtesy of [UNDP](http://hdr.undp.org/en/data). 


# Wrangling 
To merge these five different predictors into one dataset, we performed left-joins on the "Country" field. The majority of our wrangling efforts was spent on anti-joining the master dataset and the smaller one containing the predictor to reconcile different names of the same country (e.g USA vs United States of America).

### Libraries
```{r, warning = FALSE, message = FALSE}
library(readr)
library(lme4)
library(dplyr)
library(car)
```

### World Data
```{r, eval = FALSE}
# World Data
world <- map_data("world") %>%
  dplyr::select(-subregion)

# Finding countries in common that won't join because of spelling
spellingErrors <- anti_join(ecoData, world, by = c("country" = "region"))

# Fixing world names
world <- world %>%
  mutate(region = recode(region, 
                         "Antigua" = "Antigua and Barbuda",
                         "Barbuda" = "Antigua and Barbuda",
                         "Virgin Islands" = "British Virgin Islands",
                         "Brunei" = "Brunei Darussalam",
                         "Saint Kitts" = "Saint Kitts and Nevis",
                         "Saint Vincent" = "Saint Vincent and Grenadines",
                         "Trinidad" = "Trinidad and Tobago",
                         "Tobago" = "Trinidad and Tobago",
                         "UK" = "United Kingdom",
                         "USA" = "United States of America",
                         "Wallis and Futuna" = "Wallis and Futuna Islands"))

# Fixing eco names
ecoData <- ecoData %>%
  mutate(country = recode(country,
                          "Cabo Verde" = "Cape Verde",
                          "Congo" = "Republic of Congo",
                          "Congo, Democratic Republic of" = 
                            "Democratic Republic of the Congo",
                          "Côte d'Ivoire" = "Ivory Coast",
                          "Iran, Islamic Republic of" = "Iran",
                          "Korea, Democratic People's Republic of" =
                            "South Korea",
                          "Korea, Republic of" = "North Korea",
                          "Lao People's Democratic Republic" = "Laos",
                          "Libyan Arab Jamahiriya" = "Libya",
                          "Macedonia TFYR" = "Macedonia",
                          "Réunion" = "Reunion",
                          "Russian Federation" = "Russia",
                          "Syrian Arab Republic" = "Russia",
                          "Tanzania, United Republic of" = "Tanzania",
                          "Venezuela, Bolivarian Republic of" =
                            "Venezuela",
                          "Viet Nam" = "Vietnam"))

# Merging eco and world data
ecoData <- ecoData %>%
  left_join(world, by = c("country" = "region")) %>%
  dplyr::select(-region)
```

### Production Data
```{r, eval = FALSE}
### Production Data
# Reading in regional data
production <- read_csv("production.csv") %>%
  replace_na(list(region = "NA")) %>%
  group_by(Area, Year) %>%
  mutate(totalProduction = as.numeric(sum(Production))) %>%
  filter(Year == 2013) %>%
  ungroup() %>%
  distinct(Area, .keep_all = TRUE) %>%
  dplyr::select(region, Area, totalProduction) %>%
  rename("continent" = "region", "country" = "Area")

# Finding countries in common with eco Data that won't join because of spelling
spellingErrors <- anti_join(production, ecoData, by = "country")

# Fixing names and merging with production data
## Some countries are missing in eco data so there's nothing we can do about
## missing observations

## Combining China's total production
china <- production %>%
  filter(country %in% c("China, Hong Kong SAR", 
                        "China, Macao SAR",
                        "China, mainland")) %>%
  mutate(country = recode(country, 
                          "China, Hong Kong SAR" = "China",
                          "China, Macao SAR" = "China",
                          "China, mainland" = "China")) %>%
  group_by(continent, country) %>%
  summarise(totalProduction = sum(totalProduction))

## Updating production data
production <- production %>%
  filter(!(country %in% c("China, Hong Kong SAR", 
                        "China, Macao SAR",
                        "China, mainland"))) %>%
  full_join(china)

# Fixing names in production data
production <- production %>%
  mutate(country = recode(country,
                          "Bolivia (Plurinational State of)" = "Bolivia",
                          "Cabo Verde" = "Cape Verde",
                          "Congo" = "Republic of Congo",
                          "Czechia" = "Czech Republic",
                          "Democratic People's Republic of Korea" =
                            "South Korea",
                          "Iran (Islamic Republic of)" = "Iran",
                          "Lao People's Democratic Republic" = "Laos",
                          "Republic of Korea" = "North Korea",
                          "Republic of Moldova" = "Moldova",
                          "Saint Vincent and the Grenadines" =
                            "Saint Vincent and Grenadines",
                          "Russian Federation" = "Russia",
                          "The former Yugoslav Republic of Macedonia" =
                            "Macedonia",
                          "United Republic of Tanzania" = "Tanzania",
                          "Venezuela (Bolivarian Republic of)" = 
                            "Venezuela",
                          "Viet Nam" = "Vietnam"))

# Merging eco and production data 
ecoData <- ecoData %>%
  left_join(production, by = "country")
```

### Greenhouse Data
```{r, eval = FALSE}
### Greenhouse Data
# Reading in greenhouse gas dataset
greenhouse <- read_csv("CAITGHGData.csv") %>%
  filter(Year == 2012) %>%
  rename("totalGHG" = `Total GHG Emissions Including Land-Use Change and Forestry (MtCO?e?)`) %>%
  dplyr::select(Country, totalGHG)
  
# Finding countries in common that won't join because of spelling
spellingErrors <- anti_join(greenhouse, ecoData, by = c("Country" = "country"))

# Fixing names and merging with production data
## Some countries are missing in eco data so there's nothing we can do about
## missing observations
greenhouse <- greenhouse %>%
  mutate(Country = recode(Country,
                          "Antigua & Barbuda" = "Antigua and Barbuda",
                          "Bahamas, The" = "Bahamas",
                          "Bosnia & Herzegovina" = "Bosnia and Herzegovina",
                          "Brunei" = "Brunei Darussalam",
                          "Congo, Dem. Rep." = 
                            "Democratic Republic of the Congo",
                          "Congo, Rep." = "Republic of Congo",
                          "Cote d'Ivoire" = "Ivory Coast",
                          "Korea, Rep. (South)" = "South Korea",
                          "Korea, Dem. Rep. (North)" = "North Korea",
                          "Macedonia, FYR" = "Macedonia",
                          "Russian Federation" = "Russia",
                          "Saint Kitts & Nevis" = "Saint Kitts and Nevis",
                          "Saint Vincent & Grenadines" = 
                            "Saint Vincent and Grenadines",
                          "Trinidad & Tobago" = "Trinidad and Tobago",
                          "United States" = "United States of America",
                          "United States of America" = "USA"))

# Merging greenhouse gas dataset with eco dataset
ecoData <- ecoData %>%
  left_join(greenhouse, by = c("country" = "Country"))
```

### HDI Data
```{r, eval = FALSE}
# Reading in HDI data
HDI <- read_csv("HDI.csv") %>%
  rename("country" = "Country",
         "HDI2015" = `2015`) %>%
  dplyr::select(country, HDI2015)

# Finding countries in common that won't join because of spelling
spellingErrors <- anti_join(HDI, ecoData, by = "country")

# Fixing spelling errors
HDI <- HDI %>%
  mutate(country = recode(country,
         "Bolivia (Plurinational State of)" = "Bolivia",
         "Cabo Verde" = "Cape Verde",
         "Congo" = "Republic of Congo",
         "Congo (Democratic Republic of the)" = 
           "Democratic Republic of the Congo",
         "C<99>te d'Ivoire" = "Ivory Coast",
         "Iran (Islamic Republic of)" = "Iran", 
         "Korea (Republic of)" = "North Korea",
         "Lao People's Democratic Republic" = "Laos",
         "Moldova (Republic of)" = "Moldova",
         "Russian Federation" = "Russia",
         "Saint Vincent and the Grenadines" = 
           "Saint Vincent and Grenadines", 
         "Tanzania (United Republic of)" = "Tanzania",
         "The former Yugoslav Republic of Macedonia" = "Macedonia",
         "United States" = "United States of America",
         "Venezuela (Bolivarian Republic of)" = "Venezuela",
         "Viet Nam" = "Vietnam"))
HDI$country[HDI$HDI2015 == 0.474] <- "Ivory Coast"

# Merging with HDI data
ecoData <- ecoData %>%
  left_join(HDI, by = "country") %>%
  drop_na() %>%
  mutate(binomFootprint = as.factor(ifelse(netFootprint < 0, 0, 1))) %>%
  mutate(transFootprint = log(netFootprint + 15))

# eco data without world map data contained
## This form of data is better for building our models
ecoDataReg <- ecoData %>%
  distinct(country, .keep_all = TRUE) %>%
  dplyr::select(-lat, -long, -group, -order)

# Saving our final eco datasets
write_csv(ecoData, "ecoData.csv")
write_csv(ecoDataReg, "ecoDataReg.csv")
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(readr)
library(dplyr)
ecoDataReg <- read_csv("~/Documents/Swarthmore/4. Senior Year/Spring 2018/STAT 041/Stat041_p4_grp12/ecoDataReg.csv")

netFootprint <- ecoDataReg$netFootprint
transFootprint <- ecoDataReg$transFootprint
binomFootprint <- ecoDataReg$binomFootprint
population <- ecoDataReg$population
totalProduction <- ecoDataReg$totalProduction
GDPperCapita <- ecoDataReg$GDPperCapita
totalGHG <- ecoDataReg$totalGHG
HDI2015 <- ecoDataReg$HDI2015

glimpse(ecoDataReg)
```


# Model Assumptions
Before we can build our regression model, it's crucial that we first check that we haven't violated assumptions. First, let's take a look at our $y$ variable:
```{r}
# Frequency of Net Footprint 
hist(netFootprint, main = "Histogram of Net Footprint", xlab = "Net Footprint")
summary(netFootprint)
```
We see that our $y$ variable of net footprint doesn't seem to follow a normal distribution. It's skewed to the right. What we can do to fix this is take a transformation of our $y$. Let $$\text{transFootprint} = log(\text{netFootprint} + 15)$$

```{r}
# Frequency of Transformed Net Footprint
hist(transFootprint, main = "Histogram of Transformed Net Footprint",
     xlab = "Transformed Net Footprint")
```

Now let's check if our $x_i$'s and transformed $y$ are linearly related.
```{r}
# Untransformed X
plot(population, transFootprint, 
     main = "Population vs. Transformed Net Footprint",
     xlab = "Population", ylab = "Transformed Net Footprint")

# Transformed X
plot(log(population), transFootprint,
     main = "Transformed Population vs. Transformed Net Footprint",
     xlab = "log(Population)", ylab = "Transformed Net Footprint")
```

After checking the linear relationship between all of our $x_i$s and transformed $y$, we end up with $$log(y_{i} + 15) = \beta_{0} + \beta_{1}log(x_{1}) + \beta_{2}log(x_{2}) + \beta_{3}log(x_{3}) + \beta_{4}log(x_{4}) + \beta_{5}log(x_{5})$$
where $x_1$ denotes population, $x_2$ denotes GDP per capita, $x_3$ denotes total food production, $x_4$ denotes greenhouse gas emission, and $x_5$ denotes the human development index. 

Now let's check our residual and QQ plots for this model. 
```{r, warning = FALSE, message = FALSE}
# Regression 
fit1 <- lm(transFootprint ~ log(population) + log(totalProduction) +
             log(GDPperCapita) + log(totalGHG) + log(HDI2015), 
           data = ecoDataReg)

# Residual and QQ Plots
plot(fit1, which = c(1,2))
```


# Visualizations

# Hierarchical Models
### Normal Normal Hierarchical Model
```{r, warning = FALSE, message = FALSE}
normalNormal <- lmer(transFootprint ~ (1|continent) + 
                       log(population) + 
                       log(totalProduction) +
                       log(GDPperCapita) + 
                       log(totalGHG) + log(HDI2015), 
                     data = ecoDataReg)
summary(normalNormal)
```

### Binomial Normal Hierarchical Model
```{r, warning = FALSE, message = FALSE}
binomialNormal <- glmer(binomFootprint ~ (1|continent) + 
                       log(population) + 
                       log(totalProduction) +
                       log(GDPperCapita) + 
                       log(totalProduction) + 
                       log(totalGHG) + log(HDI2015), 
                     data = ecoDataReg)
summary(binomialNormal)
```


# Conclusion
